% ACM Biography for Yamil Rodríguez Montaña
% Two versions: Short (for paper) and Extended (for proceedings)

%==============================================================================
% SHORT VERSION (for ACM paper footer)
%==============================================================================

\begin{biography}{Yamil Rodríguez Montaña}
Yamil Rodríguez Montaña is Founder and Managing Partner at RefleXio, an AI venture studio focused on HealthTech and regulated AI systems. He has 15+ years of experience leading enterprise-scale data and AI transformation programs across healthcare, energy, banking, and public sector organizations. Previously, he served as Data \& AI Engineering Manager at Accenture, where he led AI-powered diagnostic platforms achieving 90\% accuracy improvements in healthcare, and as Big Data \& AI Delivery Lead at Nextret, where he directed a €6M digital transformation program serving 10M+ users for Spain's national broadcaster. His research interests include AI governance frameworks, safety-critical AI systems, and semantic validation for agentic AI in regulated industries. He holds degrees in Telecommunications Engineering and Audiovisual Communication. ACM Member 7748927.
\end{biography}

%==============================================================================
% EXTENDED VERSION (for conference proceedings or website)
%==============================================================================

\begin{biography}{Yamil Rodríguez Montaña}
Yamil Rodríguez Montaña is Founder and Managing Partner at RefleXio, an AI venture studio based in Barcelona, Spain, where he advises startups on scalable AI architectures and provides executive-level data strategy for regulated industries.

With over 15 years of experience in enterprise data and AI transformation, Yamil has led large-scale programs across healthcare, energy, banking, pharmaceuticals, media, and public sector. His work includes developing AI-powered diagnostic platforms that improved cancer detection accuracy by 90\% in public healthcare systems, architecting multi-cloud data platforms that reduced operational costs by 60\% in pharma logistics, and deploying computer vision systems that accelerated infrastructure inspection cycles by 70\% in the energy sector.

Prior to founding RefleXio, Yamil served as Data \& AI Engineering Manager at Accenture (2021–2025), where he designed cloud-native data platforms on Azure and AWS, defined enterprise governance frameworks, and led cross-functional teams implementing AI solutions in regulated environments. Before that, he was Big Data \& AI Delivery Lead at Nextret (2018–2021), directing a €6M digital transformation program for Spain's national broadcaster that delivered ML recommendation engines serving 9–10 million monthly users.

Earlier in his career, Yamil held leadership roles including Digital Transformation Lead at Synergie Spain (reporting directly to CEO), Data Product Owner at DXC Technology building fraud detection systems for major Spanish banks, and Co-Founder \& CTO of travel tech ventures (2007–2015), where he scaled engineering teams to 20+ people and achieved successful equity exit.

His research focuses on AI governance frameworks for safety-critical systems, semantic validation mechanisms for agentic AI, and the intersection of regulatory compliance (DO-178C, ISO 42001, EU AI Act) with autonomous AI deployment. He is particularly interested in bridging the gap between probabilistic AI reasoning and deterministic execution requirements in regulated industries.

Yamil holds degrees in Telecommunications Engineering and Audiovisual Communication, and is certified as a ScrumMaster (CSM) and in AWS Well-Architected Framework. He is fluent in Spanish (native), English (C1), and Catalan.

ACM Member 7748927.
\end{biography}

%==============================================================================
% ULTRA-SHORT VERSION (for author block if needed)
%==============================================================================

Yamil Rodríguez Montaña is Founder at RefleXio, an AI venture studio. He has 15+ years leading enterprise AI transformation in regulated industries including healthcare, energy, and banking. Previously Data \& AI Engineering Manager at Accenture and Big Data Lead at Nextret. ACM Member 7748927.

%==============================================================================
% INTEGRATION NOTES
%==============================================================================

% For the ACM paper, add the SHORT VERSION biography section after the 
% Acknowledgments and before References:
%
% \begin{acks}
% We thank...
% \end{acks}
%
% \begin{biography}{Yamil Rodríguez Montaña}
% [SHORT VERSION TEXT HERE]
% \end{biography}
%
% \bibliographystyle{ACM-Reference-Format}
% \begin{thebibliography}{10}
% ...
